QPLIB: A Library of Quadratic Programming Instances

### 2016-12-18 QPLIB-3.tex
checked up to §3.1, will keep checking the rest

In this section we present all the steps performed in order to build
the new library. In \S \ref{subsec:instColl}, we describe the set of gathered
instances. In \S \ref{subsec:selection}
we present the features used to classified the instances and 
we discuss the issues concerning the format of the instances. Finally in \S \ref{subsec:final set}, we describe the selection process
used to filter the instances and we graphically present the main features of selected instances.

@@

In this section we describe the procedure we adopted to gather the
instances. In January $2014$, we issued an online call for instances
using the principal international mailing lists.
In particular, we addressed the mailing lists of the mathematical
optimization and numerical analysis communities, reaching in this way
the largest possible set of interested researchers and practitioners.
The call remained open for 10 months, during which we received a large
number of contributions of different natures. The instances
we gathered come from theoretical studies and from real-world
applications.

In addition to spontaneous contribution we analysed the other generic
libraries of instances available  on internet and containing QP
instances. Namely, we cite here the libraries from which we gathered instances:

\begin{itemize}
 \item {\tt BARON Library}
\url{http://www.minlp.com/nlp-and-minlp-test-problems}
 \item {\tt CUTEst Library}
\url{https://ccpforge.cse.rl.ac.uk/gf/project/cutest/wiki}
 \item {\tt GAMS Library}
\url{http://www.gamsworld.org/performance/performlib.htm}
 \item {\tt MacMINLP} \url{https://wiki.mcs.anl.gov/leyffer/index.php/MacMINLP}
 \item {\tt Meszaros} \url{http://www.doc.ic.ac.uk/~im/00README.QP}
 \item {\tt MINLP} \url{http://www.gamsworld.org/minlp/minlplib.htm}
 \item {\tt POLIP} \url{http://polip.zib.de/pipformat.php}
\end{itemize}

Other quadratic instances were found in online libraries devoted to
specific QP problems as
Max-Cut, Quadratic Assignment, Portfolio Optimization and several
others.

At the end of this process we gathered more than eight thousand
instances. Three fourths of them contained discrete variables, while
the remaining ones contained only continuous variables.
%
More in details, we gathered a large number of Quadratic Binary Linear
(QBL), Quadratic Continuous Quadratic (QCQ) and Quadratic General
Quadratic (QGQ) instances, i.e., $\approx 1800$ instances, $\approx
2000$ instances and $\approx 2500$ instances respectively.
We gathered $\approx 1000$ Convex General Convex (CGC) instances. We
gathered relatively few Quadratic Binary Quadratic (QBQ), Convex
Continuous Convex (CCC) and Convex Mixed Convex (CMC) instances, i.e.,
$\approx 150$ instances, $\approx 200$ instances and $\approx 200$
instances respectively. Finally we gathered only $17$ Quadratic Mixed
Linear (QML) instances.

In the call for instances no specific formats were imposed
%we did not impose a specific format
for the submissions.
%In this phase we also eliminated multiples copies of the same instance.
To evaluate the instances we decided, for practical reasons, to use
Gams as common platform for all the experiments involving commercial and non-commercial
solvers.
For this reason, we decided to translate all instances into the
Gams format (\texttt{.gms}). 
% In a preliminary phase, all the instances received were divided
%according to their format and subsequently translated.
%In \S~\ref{subsec:tools}
%%~\ref{subsec:CP_convex}
%the tools used
%to translate an instance from a given format to the \texttt{.gms} format
%%to translate an instance from and to a given format to the
%\texttt{.gms} format
%are described in more detail.\\

In addition, we have introduced
a specific format called \texttt{.qplib}. This new format is capable of describing
all the instances of the library in a sparse form.
In comparison to a more \emph{high level} format like \texttt{.gms}, the new
format presents two advantages:
it is easier to read by a self-made parser and it produces smaller files.
See the Appendix~\label{sec:format} for more details.
%\textcolor{red}{ToDo: describe \texttt{.qplib} format. Maybe in an appendix?}

% is lighter in terms of sie
%  significantly
%less memory
%in comparison and

% for this reason we used the \texttt{.gms} format as one of the format of the library. %checked if one instance was present copies of the same instance, potentially written in different formats

@@

For each instance of the starting set, we collected important 
characteristics which allowed us to classify the instances into the
QP categories described in Section~\ref{sec:QPbasic}.

As far as the variable types are concerned, we collected the following information:
\begin{itemize}
\item number of binary variables, % (\emph{\# bin}),
\item number of integer variables, % (\emph{\# int}),
\item number of continuous variables. % (\emph{\# cont}).
\end{itemize}

In case at least one binary or integer variable is present, then the
instance is categorized as a \emph{discrete} instance, on the other
hand, in case only continuous variables are presents, the instance is
categorized as a \emph{continuous} instance.

As fas as the objective function is concerned, we gathered the
following information:
\begin{itemize}
\item percentage of negative eigenvalues of the $Q^0$ matrix,  % (\emph{\% neg eig}),
\item density of the $Q^0$ matrix . %(\emph{\% dens}).
\end{itemize}
 The number of negative eigenvalues of $Q^0$ allows us to
identify the specific objective function type, i.e., in presence of at
least one negative eigenvalue the objective function is non convex.
The density of $Q^0$ is defined as the
number of nonzero entries  over the total number.

As far as the constraint types are concerned, we collected the following information:
\begin{itemize}
\item number of linear constraints, % (\emph{\# lin}),
\item number of quadratic constraints, % (\emph{\# quad}),
\item number of convex constraints, % (\emph{\# conv}),
\item number of box constraints. %(\emph{\# box}).
\end{itemize}
A constraint is considered quadratic if it contain a quadratic term.
Among the quadratic constraints, the one with only non negative
eigenvalues are classified as convex constraints.

This information allowed us to analyze the gathered instances and to perform the filters described in the the next paragraph.

\subsection{Instance Selection}\label{subsec:selection}

@@

\subsection{Analysis of the final set of instances}\label{subsec:final set}
In this section we analyse the feature of the instances selected to be
part of library.
The characteristics of the instances in the final library are
presented in Table \ref{tab:DD} for \emph{discrete} instances
(*\{B,M,I,G\}*) and in Table \ref{tab:CC} for
\emph{continuous} ones (*C*).
For each category, the Tables report in column $\#$ the corresponding
number of instances.
The final set well respects the original distribution of the gathered
instances among the different categories.
It is worth noticing that the classical discrete categories (such as
(L,M,Q) or (Q,B,L)) are well represented by $118$ and $59$  instances
respectively. On the other side, the classical continuous categories
(such as (L,C,Q) and (QCQ)) are also well represented by $50$ and $17$
 instances respectively.
Moreover, the library covers a large majority of all possible
categories of instances.

>>>>

In this section we present all the steps performed in order to build the
new library. In \S \ref{subsec:instColl}, we describe the set of gathered
instances. In \S \ref{subsec:selection} we present the features used to
classify the instances and we discuss the issues concerning the format of
the instances. Finally, in \S \ref{subsec:final set}, we describe the
selection process used to filter the instances and we graphically present
the main features of the selected instances.

@@

In this section we describe the procedure we adopted to gather the
instances. In January $2014$, we issued an online call for instances
using the main international mailing lists of the mathematical
optimization and numerical analysis communities, reaching in this way
the largest possible set of interested researchers and practitioners.
The call remained open for 10 months, during which we received a large
number of contributions of different nature. The instances we gathered
come both from theoretical studies as well as from real-world
applications.

In addition to spontaneous contribution we analysed the other generic
libraries of instances available  on internet and containing QP
instances. Namely, the libraries from which we gathered instances are
%
\begin{itemize}
 \item the {\tt BARON} library
 \url{http://www.minlp.com/nlp-and-minlp-test-problems};
%
\item the {\tt CUTEst} library
 \url{https://ccpforge.cse.rl.ac.uk/gf/project/cutest/wiki};
%
\item the {\tt GAMS} library
 \url{http://www.gamsworld.org/performance/performlib.htm};
%
\item the {\tt MacMINLP} library
 \url{https://wiki.mcs.anl.gov/leyffer/index.php/MacMINLP};
%
\item the {\tt Meszaros} library
 \url{http://www.doc.ic.ac.uk/~im/00README.QP};
%
\item the {\tt MINLP} library
 \url{http://www.gamsworld.org/minlp/minlplib.htm};
%
\item the {\tt POLIP} library
 \url{http://polip.zib.de/pipformat.php}.
\end{itemize}

Other quadratic instances were found in online libraries devoted to
specific QP problems as Max-Cut, Quadratic Assignment, Portfolio
Optimization, and several others.

At the end of this process we gathered more than eight thousand
instances. Three fourths of them contained discrete variables, while
the remaining ones contained only continuous variables. More in details,
we gathered $\approx 1800$ Quadratic Binary Linear (QBL) instances,
$\approx 2000$ Quadratic Continuous Quadratic (QCQ) instances, and
and $\approx 2500$ Quadratic General Quadratic (QGQ) instances. W
also gathered $\approx 1000$ Convex General Convex (CGC) instances. We
gathered relatively fewer Quadratic Binary Quadratic (QBQ), Convex
Continuous Convex (CCC) and Convex Mixed Convex (CMC) instances, i.e.,
$\approx 150$ instances, $\approx 200$ instances and $\approx 200$
instances respectively. Finally, we gathered only $17$ Quadratic Mixed
Linear (QML) instances. In the call for instances, no specific formats
requirements were imposed for the submissions.

To evaluate the instances we decided, for practical reasons, to use
Gams as common platform for all the experiments involving commercial and
non-commercial solvers. For this reason, we decided to translate all
instances into the Gams format (\texttt{.gms}). 
% In a preliminary phase, all the instances received were divided
%according to their format and subsequently translated.
%In \S~\ref{subsec:tools}
%%~\ref{subsec:CP_convex}
%the tools used
%to translate an instance from a given format to the \texttt{.gms} format
%%to translate an instance from and to a given format to the
%\texttt{.gms} format
%are described in more detail.\\
%
In addition, we have introduced a new, specific \texttt{.qplib} format. This
new is capable of describing all the instances of the library in a sparse form.
In comparison to a more \emph{high level} format like \texttt{.gms}, the new
format presents two advantages: it is easier to read by a self-made parser and
it produces smaller files. See the Appendix~\label{sec:format} for more details.

@@

For each instance of the starting set, we collected important characteristics
which allowed us to classify the instances into the QP categories described in
Section~\ref{sec:QPbasic}. As far as the variable types are concerned, we
collected the following information:
%
\begin{itemize}
 \item number of binary variables; % (\emph{\# bin}),
 %
 \item number of integer variables; % (\emph{\# int}),
 %
 \item number of continuous variables. % (\emph{\# cont}).
\end{itemize}
%
In case at least one binary or integer variable is present, then the instance is
categorized as \emph{discrete}, otherwise it is categorized as \emph{continuous}.
As fas as the objective function is concerned, we gathered the following
information:
%
\begin{itemize}
 \item percentage of negative eigenvalues of the $Q^0$ matrix;
       % (\emph{\% neg eig}),
 %
 \item density of the $Q^0$ matrix (number of nonzero entries over the total
       number). %(\emph{\% dens}).
\end{itemize}
%
The number of negative eigenvalues of $Q^0$ allows us to identify the
objective function type, as in presence of at least one negative eigenvalue
the objective function is non convex. Finally, as far as the constraint types
are concerned, we collected the following information:
%
\begin{itemize}
 \item number of linear constraints, % (\emph{\# lin}),
 %
 \item number of quadratic constraints, % (\emph{\# quad}),
 %
 \item number of convex constraints, % (\emph{\# conv}),
 %
 \item number of box constraints. %(\emph{\# box}).
\end{itemize}
%
A constraint is considered quadratic if it contains at least one nonzero in
the quadratic term. Among the quadratic constraints, the ones with only
non-negative eigenvalues are classified as convex constraints. All this
information allowed us to analyze the gathered instances and to perform the
filters described in the the next paragraph.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Instance Selection}\label{subsec:selection}

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Analysis of the final set of instances}\label{subsec:final set}

In this section we analyse the feature of the instances selected to be
part of library.
The characteristics of the instances in the final library are
presented in Table \ref{tab:DD} for \emph{discrete} instances
(*\{B,M,I,G\}*) and in Table \ref{tab:CC} for
\emph{continuous} ones (*C*).
For each category, the Tables report in column $\#$ the corresponding
number of instances.
The final set well respects the original distribution of the gathered
instances among the different categories.
It is worth noticing that the classical discrete categories (such as
(L,M,Q) or (Q,B,L)) are well represented by $118$ and $59$  instances
respectively. On the other side, the classical continuous categories
(such as (L,C,Q) and (QCQ)) are also well represented by $50$ and $17$
 instances respectively.
Moreover, the library covers a large majority of all possible
categories of instances.
